---
title: "Algorithms Assignment 3"
author: "Eric Folsom"
format: html
---

```{python, setup_py}
#| echo: false
#| output: false

# Setup Chunk for python, loading the required libraries
import numpy as np
import pandas as pd
```


```{r, setup_r}
#| echo: false
#| output: false

#Setup Chunk for R and Reticulate so that I can use ggplot if I ever have to graph stuff.
#This will most likely be useful for a later assignment.

library(reticulate)
library(tidyverse)
library(ggplot2)
```


## Problem 1: Write a linear search algorithm both recursively and iteratively in pseudo code. Comment to explain.

## Problem 2: Write the same linear search algorithms in the programming language of your choice. Comment to explain.

## Problem 3: Rewrite the binary search algorithm from week 1 in a recursive form in pseudo code and in the programming language of your choice. Comment to explain.

## Problem 4: Find a for loop you wrote in previous course work and rewrite it as a recursive function (it can not be a simple linear search problem). Please show the original code in your submission.

Dig through some R files on your computer at home. Then make sure it works within this script, then change it to be recursive, make it simple as some of the R stuff will run for a long ass time.

```{r, original iterative function}
x<-rnorm(1000)
df<- data.frame(x)
ggplot(df, aes(x=x))+geom_density()


```


```{r, recursive version}

```

## Problem 5: 

