---
title: "Algorithms Assignment 3"
author: "Eric Folsom"
format: html
---

```{python, setup_py}
#| echo: false
#| output: false

# Setup Chunk for python, loading the required libraries
import numpy as np
import pandas as pd
```


```{r, setup_r}
#| echo: false
#| output: false

#Setup Chunk for R and Reticulate so that I can use ggplot if I ever have to graph stuff.
#This will most likely be useful for a later assignment.

library(reticulate)
library(tidyverse)
library(ggplot2)
```


## Problem 1: Write a linear search algorithm both recursively and iteratively in pseudo code. Comment to explain.

#### Iteratively:

For an iterative linear search, the pseudocode is as follows:

We start with an array, $X = \{x_0, x_1, x_2, \ldots, x_n\}$. Where $x_i, i = 0,1,2,\ldots,n$ are the elements of the array.

Suppose we are searching for the element in $X$, $x_j$. We would impelement something similar to the following:

1. Start at $x_0$. Keep track of what index we are searching in.
    * Compare the value of $x_0$ with the value $x_j$.
      * If $x_j$ = x_0:
        * Return the index that $x_j$ is in
      * Else:
        * 


#### Recursively:

## Problem 2: Write the same linear search algorithms in the programming language of your choice. Comment to explain.

I will be implementing these linear serach algorithms in both R and python. 

The linear search algorithm in R is as follows:

First we will do it iteratively:

```{r, linear search iterative}

```

Next we will do it recursively:

```{r, linear serach recursive}

```

Next we will implement these algorithms in python.

We begin with the iterative implementation:

```{python, linear search iterative}


```

Next we do it recursively:

```{python, linear serach recursive}

```

## Problem 3: Rewrite the binary search algorithm from week 1 in a recursive form in pseudo code and in the programming language of your choice. Comment to explain.

## Problem 4: Find a for loop you wrote in previous course work and rewrite it as a recursive function (it can not be a simple linear search problem). Please show the original code in your submission.

Dig through some R files on your computer at home. Then make sure it works within this script, then change it to be recursive, make it simple as some of the R stuff will run for a long ass time. Maybe do the kmeans choice algorithm for your 447 project?

```{r, original iterative function}
x<-rnorm(1000)
df<- data.frame(x)
ggplot(df, aes(x=x))+geom_density()


```


```{r, recursive version}

```

## Problem 5: 

