---
title: "CS240 Data Structures and Algorithm Fundametals Notes"
format: html
---
```{r, setup_R}
#| echo: false
#| output: false

library(tidyverse)
library(reshape2)

```

```{python, setup_py}
#| echo: false
#| output: false

import numpy as np
import pandas as pd
```

# General Information

* Instructor: Darrel Criss
        
    Office: Baker Hall 123

* Class meets from 3:30-4:50, Monday, Tuesday, Wednesday. Baker Hall 115

* Office Hours: By Appointment or Thursdays from 12:00pm-1:00pm.

* Homeworks will be due Tuesdays at midnight

# Course Notes

### April 2, 2024

We wnet over the basics of the course as well as talked about rules for Using ChatGPT. 

### April 3, 2024 

Again went over the basics and the course structure. A little bit of repetition as some of the class was not there on Tuesday April 2. Talked with Prof after class to ask about graduate certifficate vs Graduate degree as well as talked a little about AI.

### April 4, 2024

Lab Day, Going to work on homework and figuring out how python works. Hoping to transition to quarto vs. Rmarkdown.

* Standups
  - A chance to show off some of your code
  - Show off your aproach to solving the problem / build the prject
  - technical description of the solution
  - What we are happy with
  - What you would like to improve later

* Code Review
    - Be Polite and professional
    - Review of syntax
    - Evaluation of any edge cases missed
    - Notes if you approached it in a different way, used a different library

* To DO: 
   - Finish introductions - include a picture
   - Finish Readings
   - Finish Algorithms Assignment 1

###### Algorithms Assignment 1

Binary Search and Time Complexity

For the first question, we are writing pseudocode to write an alogrithm for performing the binary search across an Array.

* Small Note on O-Notation:
  
   - $O(1)$ would be an example of an array like adding a value to an array, something in memory that is unchanging. Fairly Fast.
   - $O(log_2(n))$ Is an example of Binary Search Trees. Fairly fast, as n increases, gets large slower and slower. 
   - $O(n^2)$ Gets slow very fast.
  
* In more general:
  
  |$\#$ Items to search | $\#$ Number searches using BST|
  |:------------------:|:----------------------------:|
  |1 K|10|
  |1Meg|20|
  |1Gig|30|
  |1 Tera| 40|

Big Oh Notation: $O(n)$ where $n$ usually refers to the size of the input or the number of elements in the input. The function $O(f(n))$ tells us the number of operations that some function of the number of elements $n$, $f(n)$ will take. This is not very mathematically rigorous because we usually drop constants in front of $n$ (number of elements). Computation time.

|Name|Notation|
|:-:|:-:|
|Linear Time|$O(n)$|
|Constant Time| $O(1)$|
|Quadratic Time| $O(n^2)$|

* How to find the Time complexity of a given function of $n$:
  1. Find the fastest growing term.
  2. Take out the coefficeint.

$T=an+b = O(n)$, here $an>b$ because $b$ is constant.

$T=cn^2+dn+e=O(n^2)$, here $cn^2>dn>e$ (graph them itll make more sense)

```{r, graph visualization thing}
#| echo: false


x<-c(1:5)
y<-x
y2<-x^2
con<-rep(5,5)
df<-data.frame(x,y,y2,con)

df<-melt(df,id.vars="x")
ggplot(df, aes(x=x, y= value, color= variable))+geom_line()
```


### April 8, 2024



```{r}

```