---
title: "CS240 Data Structure and Algorithm Fundamentals"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r include=FALSE}
library(reticulate)
```

## Using Python inside of Rmarkdown

Create a variable `x` in the Python Session:

```{python echo=TRUE}
x = [1, 2, 3]
```

Access the Python Variable `x` in an R code chink"

```{r echo=TRUE}
py$x
```

Create a new Variable `y` in the Python Session using R, and pass a data frame to `y`:

```{r echo=TRUE}
py$y<-head(cars)
```

Print the variable `y` in Python:

```{python echo=TRUE}
print(y)
```

# Course Notes

### April 2, 2024

I was a few min. late :/


##### Course Introduction:

cs141 problem solving with java

cs145 intro to data structures - efficiency and O notation
 
THis course:


Chat GPT Rules: 

  YOu are responsible for your output
  
  understand how it works
  
  cite sources - includes chatgpt
  
  
  
Gonna use visual studio, gotta get it and get my rmd files running.  



